To my advisor, Sami Assaf---thank you not just for your support and excitement
about all of mathematical interests (recreational and otherwise), but also for your
friendship, encouragement, and generosity since even before I started at USC.
My time here has been a joy, and you're largely to blame for that.
The way that you supported for me and advocated for me is
an inspiration for how I want to treat my students.

To my committee member, Richard Arratia---the sort of chance encounters I
regularly had with you is what I dreamed about before coming to graduate school.
You've always been generous to me with your time, with your ideas, with your
bounties, with your jokes, and with your great taste in problems.
It was a real pleasure getting to TA for you and get to know you these past five years.
I'm really going to miss having my office across the hall from yours.

To my committee member, David Kempe---It has been a blessing having your
thoughtfulness and interest throughout this process.
I've left every interaction with you full of new ideas and possible directions,
and I hope our paths continue to cross.

I would like to thank all of my colleagues in the program, especially those
teamed up to do help out with the important work in KAP 500 week after week.
I'd like to acknowledge the support that I got from my family and friends
outside of the program. It's a blessing to have such an engaged and caring
group of folks supporting me.

Finally, Sierra---getting to spend half of a decade with you in graduate school
has been the blessing of a lifetime. You make the world a richer place, and I
couldn't be more thankful to have you in my corner.
You inspire me as a student, as an educator, and as a person.
I can't wait to see what the rest of our lives have in store together!
